---
title: "ABDWRChapter8Supp"
output: html_document
---

## Baseball OPS and Peak Ages

This supplement is just to take a look at using the `brms` package to fit a hierarchical model of OPS and peak ages for baseball players. The data is from the `Lahman` package, which contains historical baseball statistics. This is a supplement to my notes on Chapter 8.


### Grab all the data. 
This is a subset of the code from the book just to get the data ready for analysis.
```{r}
library(tidyverse)
library(Lahman)
library(brms)

batting <- Batting |>
  replace_na(list(SF = 0, HBP = 0))

get_stats <- function(player_id) {
  batting |> 
    filter(playerID == player_id) |>
    inner_join(People, by = "playerID") |>
    mutate(
      birthyear = if_else(
        birthMonth >= 7, birthYear + 1, birthYear
      ),
      Age = yearID - birthyear,
      SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB,
      OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
      OPS = SLG + OBP
    ) |>
    select(Age, SLG, OBP, OPS)
}

batting_2000 <- batting |> 
  group_by(playerID) |>
  summarize(AB_career = sum(AB, na.rm = TRUE)) |>
  inner_join(batting, by = "playerID") |>
  filter(AB_career >= 2000)


Positions <- Fielding |> 
  group_by(playerID, POS) |>
  summarize(Games = sum(G)) |> 
  arrange(playerID, desc(Games)) |> 
  filter(POS == first(POS))

batting_2000 <- batting_2000 |>
  inner_join(Positions, by = "playerID")


batting_2000 <- batting_2000 |> 
  group_by(playerID, yearID) |>
  summarize(
    G = sum(G), AB = sum(AB), R = sum(R),
    H = sum(H), X2B = sum(X2B), X3B = sum(X3B),
    HR = sum(HR), RBI = sum(RBI), SB = sum(SB),
    CS = sum(CS), BB = sum(BB), SH = sum(SH),
    SF = sum(SF), HBP = sum(HBP),
    AB_career = first(AB_career),
    POS = first(POS)
  ) |>
  mutate(
    SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB,
    OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
    OPS = SLG + OBP
  )
batting_2000 <- batting_2000 |>
  inner_join(People, by = "playerID") |>
  mutate(
    Birthyear = if_else(
      birthMonth >= 7, birthYear + 1, birthYear
    ),
    Age = yearID - Birthyear
  )

batting_2000 |> drop_na(Age) -> batting_2000


not_current_playerID <- People |>
  filter(finalGame < "2021-11-01") |> 
  pull(playerID)

batting_2000 <- batting_2000 |>
  filter(playerID %in% not_current_playerID)


midcareers <- batting_2000 |>
  group_by(playerID) |>
  summarize(
    Midyear = (min(yearID) + max(yearID)) / 2,
    AB_total = first(AB_career)
  )
batting_2000 <- batting_2000 |>
  inner_join(midcareers, by = "playerID")

# Add fielding positions
batting_2000a <- batting_2000 |>
  filter(Midyear >= 1985, Midyear <= 1995) |> ungroup() 

batting_2000a <- batting_2000a |>
  mutate(age_c = (Age - 30)/20)|>   # rescaled and centered-ish
  select(OPS, playerID, Age, age_c, POS) |> drop_na()

```


Idea here is to use the playerID as 'group' to share information across players, and then use the age as a predictor.  The quadratic term is to allow for the peak age to be estimated.  We use a student - t distribution to allow for the heavy tails (posterior predictive check without it looked pretty bad)


```{r}
fit_player <- brm(
  OPS ~ age_c + I(age_c^2) + (age_c + I(age_c^2) | playerID),
  data = batting_2000a,
  cores = 4, iter = 2000,
  family = 'student',
  control = list(adapt_delta = 0.95)
)
```

```{r}

# Posterior predictive check
pp_check(fit_player)
```



```{r}
posterior_player <- as_draws_df(fit_player)
# Compute peak ages per player

player_peak_ages <- posterior_player |>
  select(starts_with("r_playerID"), b_age_c, b_Iage_cE2) |>
  pivot_longer(cols = starts_with("r_playerID"),
               names_to = c("playerID", "param"),
               names_pattern = "r_playerID\\[(.*?),(.*?)\\]") |>
  pivot_wider(names_from = param, values_from = value) |>
  mutate(
    linear_total = b_age_c + age_c,
    quad_total = b_Iage_cE2 + Iage_cE2,
    peak_age = 30 - 10*(linear_total / (2 * quad_total))
  ) %>%
  group_by(playerID) %>%
  summarize(
    peak_age_median = median(peak_age),
    peak_age_lower = quantile(peak_age, 0.25),
    peak_age_upper = quantile(peak_age, 0.75)
  )  

player_peak_ages <- player_peak_ages %>%
  left_join(batting_2000a %>% select(playerID, POS) %>% distinct(), by = "playerID")
```

```{r}
# Plot strip plot grouped by position
ggplot(player_peak_ages, aes(x = POS, y = peak_age_median)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  #geom_errorbar(
  #  aes(ymin = peak_age_lower, ymax = peak_age_upper),
  #  width = 0.1, alpha = 0.4, color = "gray40"
  #) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  labs(title = "Player Peak Ages Grouped by Position", 
       x = "Position", y = "Peak Age") +
  #scale_y_continuous(limits = c(20,50)) +
  theme_minimal()
```

With error bars 

```{r}
# jitter by hand
player_peak_ages <- player_peak_ages %>%
  mutate(POS_jitter = as.numeric(factor(POS)) + runif(n(), -0.15, 0.15))

ggplot(player_peak_ages, aes(x = POS_jitter, y = peak_age_median)) +
  geom_point(alpha = 0.6, color = "blue", size = 1.5) +
  geom_errorbar(
    aes(ymin = peak_age_lower, ymax = peak_age_upper),
    width = 0.0, alpha = 0.4, color = "gray40"
  ) +
  scale_x_continuous(
    breaks = 1:length(unique(player_peak_ages$POS)),
    labels = sort(unique(player_peak_ages$POS))
  ) +
 
  labs(title = "Player Peak Ages Grouped by Position (with uncertainty)", 
       x = "Position", y = "Peak Age") +
  theme_minimal()
```


We can also check the posterior distribution of fit curves.

```{r}
# Select 4 random players
set.seed(123)
selected_players <- sample(unique(batting_2000a$playerID), 4)

 
age_seq <- seq(20, 50, length.out = 50)
plot_data <- expand.grid(playerID = selected_players, Age = age_seq) %>%
  mutate(age_c = (Age - 30)/20)

# Get linear predictor predictions (30 draws)
linpred_matrix <- posterior_linpred(fit_player, newdata = plot_data, ndraws = 30)

 
pred_df <- plot_data %>%
  mutate(row = row_number()) %>%
  left_join(
    as.data.frame(t(linpred_matrix)) %>%
      mutate(row = row_number()) %>%
      pivot_longer(-row, names_to = "draw", values_to = "OPS_pred"),
    by = "row"
  )

 
actual_df <- batting_2000a %>%
  filter(playerID %in% selected_players)

 
ggplot() +
  geom_point(data = actual_df, aes(x = Age, y = OPS), color = "black") +
  geom_line(data = pred_df, aes(x = Age, y = OPS_pred, group = draw), 
            color = "blue", alpha = 0.3) +
  facet_wrap(~playerID, scales = "free") +
  labs(title = "Smooth Fitted OPS curves (30 posterior draws) vs. Actual Data",
       x = "Age",
       y = "OPS") +
  theme_minimal()

```

## Nonlinear 

We note that the trajectories that are allowed can be quite flexible, in fact too flexible, allowing for trajectories that grow with time for ever (i.e. the quadratic term is positive).  We can use a nonlinear parameterization to force the peak age to be stable.  Fortunately, `brms` allows us to do this easily with the `nl` argument in the `bf()` function.  We can also use informative priors to help guide the model to a reasonable solution.

Note that this model still allows an individual players gamma (the curvature of the trajectory) to be negative, because in some cases the trajectory really might be concave up.    

```{r}
nlform <- bf(
  OPS ~ peakOPS - gamma * ((Age - peakAge)/10)^2,
  peakOPS + peakAge + gamma ~ 1 + (1 |ID| playerID),
  nl = TRUE
)

# Set informative priors
priors <- c(
  prior(normal(0.75, 0.2), nlpar = "peakOPS", lb = 0.5, ub = 2),  # peak OPS
  prior(normal(30, 5), nlpar = "peakAge", lb = 10, ub = 60),  # restrict peak age
  prior(exponential(1), nlpar = "gamma", lb = 0)  # gamma should be positive
)

fit_peak <- brm(
  nlform,
  data = batting_2000a,
  prior = priors,
  cores = 4,
  iter = 4000,
  family = student(),
  control = list(adapt_delta = 0.95)
)
```

```{r}
pp_check(fit_peak)
```


Plot 

```{r}
# Select 4 random players (use same players as earlier for consistency)
set.seed(123)
selected_players <- sample(unique(batting_2000a$playerID), 4)
 
age_seq <- seq(min(batting_2000a$Age), max(batting_2000a$Age), length.out = 50)
plot_data <- expand.grid(playerID = selected_players, Age = age_seq)

epreds <- posterior_epred(fit_peak, newdata = plot_data, ndraws = 30)

pred_df <- plot_data %>%
  mutate(row = row_number()) %>%
  left_join(
    as.data.frame(t(epreds)) %>%
      mutate(row = row_number()) %>%
      pivot_longer(-row, names_to = "draw", values_to = "OPS_pred"),
    by = "row"
  )

 
actual_df <- batting_2000a %>%
  filter(playerID %in% selected_players)
 
ggplot() +
  geom_point(data = actual_df, aes(x = Age, y = OPS), color = "black") +
  geom_line(data = pred_df, aes(x = Age, y = OPS_pred, group = draw),
            color = "blue", alpha = 0.3) +
  facet_wrap(~playerID, scales = "free") +
  labs(title = "Posterior OPS predictions vs. Actual Data",
       subtitle = "30 posterior draws using stable peak-age parameterization",
       x = "Age",
       y = "OPS") +
  theme_minimal()

```

If time permitted we could explore this further. For example, in this model we can now include directly POS as a predictor that we can interprets for peak age.  I could not get this to sample well yet, need to rethink this sometime in the future.  

```{r}
nlformops <- bf(
  OPS ~ peakOPS - gamma * ((Age - peakAge)/10)^2,
  gamma ~ (1 | ID | playerID),
  peakAge + peakOPS ~ 0 + POS + (1 | ID |playerID),
  nl = TRUE
)

# Set informative priors
priors <- c(
  prior(normal(0.75, 0.2), nlpar = "peakOPS", lb = 0.5, ub = 2),  # peak OPS
  prior(normal(30, 5), nlpar = "peakAge", lb = 10, ub = 60),  # restrict peak age
  prior(exponential(1), nlpar = "gamma", lb = 0)  # gamma should be positive
)

fit_peakops <- brm(
  nlformops,
  data = batting_2000a,
  prior = priors,
  cores = 4,
  iter = 4000,
  family = student(),
  control = list(adapt_delta = 0.95)
)
```



## Fake data validation


Simple test of hierarchical peak model


Fake data:
```{r}
library(brms)
library(dplyr)
library(tidyverse)

# Set seed
set.seed(123)

 
n_positions <- 8
positions <- paste0("Pos", 1:n_positions)
n_players <- 80

# True position-specific parameters
true_intercepts <- rnorm(n_positions, 0.75, 0.05)
true_linear_age <- rnorm(n_positions, 0.01, 0.005)
true_quad_age <- rnorm(n_positions, -0.002, 0.0005)

# Expand each player across ages 20 to 40
fake_data <- expand.grid(
  player = 1:n_players,
  age = 20:40
) %>%
  mutate(
    position = sample(positions, n_players, replace = TRUE)[player],
    pos_idx = match(position, positions),
    age_c = age - 30,
    ops = true_intercepts[pos_idx] + 
          true_linear_age[pos_idx] * age_c + 
          true_quad_age[pos_idx] * age_c^2 + 
          rnorm(n(), 0, 0.02)
  )
```

Example

```{r}
random_player <- sample(unique(fake_data$player), 1)
fake_data %>%
  filter(player == random_player) %>%
  ggplot(aes(x = age, y = ops)) +
  geom_point() +
  geom_line() +
  labs(title = paste("OPS vs Age for Player", random_player),
       x = "Age", y = "OPS")
```


fit model

```{r}
# Fit hierarchical model
fit_player <- brm(
  ops ~ age_c + I(age_c^2) + (age_c + I(age_c^2) | player),
  data = fake_data,
  cores = 4, iter = 2000,
  control = list(adapt_delta = 0.95)
)
summary(fit)

# Posterior predictive check
pp_check(fit)
```


Compute each player peak ages

```{r}

posterior_player <- as_draws_df(fit_player)

# Compute peak ages per player
player_peak_ages <- posterior_player %>%
  select(starts_with("r_player"), b_age_c, b_Iage_cE2) %>%
  pivot_longer(cols = starts_with("r_player"), names_to = c("player", "param"),
               names_pattern = "r_player\\[(.*?),(.*?)\\]") %>%
  pivot_wider(names_from = param, values_from = value) %>%
  mutate(
    linear_total = b_age_c + age_c,
    quad_total = b_Iage_cE2 + Iage_cE2,
    peak_age = 30 - (linear_total / (2 * quad_total))
  ) %>%
  group_by(player) %>%
  summarize(
    peak_age_mean = mean(peak_age),
    peak_age_lower = quantile(peak_age, 0.025),
    peak_age_upper = quantile(peak_age, 0.975)
  ) %>%
  mutate(player = as.integer(player))

```


And plot
```{r}
player_peak_ages <- player_peak_ages %>%
  left_join(fake_data %>% select(player, position) %>% distinct(), by = "player")

# Plot strip plot grouped by position
ggplot(player_peak_ages, aes(x = position, y = peak_age_mean)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  labs(title = "Player Peak Ages Grouped by Position", 
       x = "Position", y = "Peak Age") +
  theme_minimal()
``` 


## Even more complicated model

```{r}
fit_player <- brm(
  ops ~ age_c + I(age_c^2) + 
    (age_c  | position) + 
    (age_c + I(age_c^2) | player),
  data = fake_data,
  cores = 4, iter = 3000,
  control = list(adapt_delta = 0.95)
)
```
