---
title: "ABDWRChapter9"
output: html_document
---

## This chapter is about "Simulation"

```{r}
library(tidyverse)
library(abdwr3edata)
retro2016 <- read_rds(here::here("data/retro2016.rds"))
```


## Compute transition matrix

Add states and generate transition matrix.

```{r}
retro2016 <- retro2016 |>
  retrosheet_add_states()

half_innings <- retro2016 |>
  mutate(
    runs = away_score_ct + home_score_ct,
    half_inning_id = paste(game_id, inn_ct, bat_home_id)
  ) |>
  group_by(half_inning_id) |>
  summarize(
    outs_inning = sum(event_outs_ct), 
    runs_inning = sum(runs_scored),
    runs_start = first(runs),
    max_runs = runs_inning + runs_start
  )

retro2016_complete <- retro2016 |> 
  mutate(
    half_inning_id = paste(game_id, inn_ct, bat_home_id)
  ) |>
  inner_join(half_innings, join_by(half_inning_id)) |>
  filter(state != new_state | runs_scored > 0) |> 
  filter(outs_inning == 3, bat_event_fl)

retro2016_complete <- retro2016_complete |>
  mutate(new_state = str_replace(new_state, "[0-1]{3} 3", "3"))

T_matrix <- retro2016_complete |>
  select(state, new_state) |>
  table()


P_matrix <- prop.table(T_matrix, 1)  |>
  rbind("3" = c(rep(0, 24), 1))
```

Example


```{r}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0)
```

## Simulate markov chain

```{r}
num_havent_scored <- function(s) {
  s |>
    str_split("") |>
    pluck(1) |>
    as.numeric() |>
    sum(na.rm = TRUE)
}

runners_out <- T_matrix |>
  row.names() |>
  set_names() |>
  map_int(num_havent_scored)

# Runs for each transition
R_runs <- outer(
  runners_out + 1, 
  runners_out, 
  FUN = "-"
) |>
  cbind("3" = rep(0, 24))

simulate_half_inning <- function(P, R, start = 1) {
  s <- start
  path <- NULL
  runs <- 0
  while (s < 25) {
    s_new <- sample(1:25, size = 1, prob = P[s, ])
    path <- c(path, s_new)
    runs <- runs + R[s, s_new]
    s <- s_new
  }
  runs
}

```

Example

```{r}
set.seed(111653)
simulated_runs <- 1:10000 |>
  map_int(~simulate_half_inning(T_matrix, R_runs))

table(simulated_runs)
```


## Runs potential

```{r}
runs_j <- function(j) {
  1:10000 |>
    map_int(~simulate_half_inning(T_matrix, R_runs, j)) |>
    mean()
}

erm_2016_mc <- tibble(
  state = row.names(T_matrix), 
  mean_run_value = map_dbl(1:24, runs_j)
) |>
  mutate(
    bases = str_sub(state, 1, 3),
    outs_ct = as.numeric(str_sub(state, 5, 5))
  ) |>
  select(-state)
  
erm_2016_mc |>
  pivot_wider(names_from = outs_ct, values_from = mean_run_value)

```

This can be compared to previously computed erm2016.rds, they are different due to the plays we left out (e.g. steals)

## Simulating a baseball season (Bradley Terry)

Set up
```{r}
make_schedule <- function(teams, k) {
  num_teams <- length(teams)
  Home <- rep(rep(teams, each = num_teams), k)
  Visitor <- rep(rep(teams, num_teams), k)
  tibble(Home = Home, Visitor = Visitor) |>
    filter(Home != Visitor)
}

library(Lahman)
teams_68 <- Teams |>
  filter(yearID == 1968) |>
  select(teamID, lgID) |>
  mutate(teamID = as.character(teamID)) |>
  group_by(lgID)

schedule <- teams_68 |>
  group_split() |>
  set_names(pull(group_keys(teams_68), "lgID")) |>
  map(~make_schedule(teams = .x$teamID, k = 9)) |>
  list_rbind(names_to = "lgID")

# Simulated talent from normal distribution
s_talent <- 0.20
teams_68 <- teams_68 |>
  mutate(talent = rnorm(10, 0, s_talent))

schedule_talent <- schedule |>
  inner_join(teams_68, join_by(lgID, Home == teamID)) |>
  rename(talent_home = talent) |>
  inner_join(teams_68, join_by(lgID, Visitor == teamID)) |>
  rename(talent_visitor = talent)



schedule_talent <- schedule_talent |> 
  mutate(
    prob_home = exp(talent_home) /
      (exp(talent_home) + exp(talent_visitor))
  )

# THe actual simulation
schedule_talent <- schedule_talent |>
  mutate(
    outcome = rbinom(nrow(schedule_talent), 1, prob_home),
    winner = if_else(outcome == 1, Home, Visitor)
)

schedule_talent |>
  select(Visitor, Home, prob_home, outcome, winner) |> 
  slice_head(n = 6)
```


```{r}
results <- schedule_talent |> 
  group_by(winner) |>
  summarize(Wins = n()) |>
  inner_join(teams_68, by = c("winner" = "teamID"))
results
```

## Simulating the post season

```{r}
win_league <- function(res) {
  res |>
    group_by(lgID) |>
    mutate(
      tiebreaker = runif(n = length(talent)),
      wins_total = Wins + tiebreaker,
      rank = min_rank(desc(wins_total)),
      is_winner_lg = wins_total == max(wins_total)
    )
}

sim_one <- win_league(results)

ws_winner <- sim_one |>
  filter(is_winner_lg) |>
  ungroup() |>
  mutate(
    outcome = as.numeric(rmultinom(1, 7, exp(talent))),
    is_winner_ws = outcome > 3
  ) |>
  filter(is_winner_ws) |>
  select(winner, is_winner_ws)

sim_one |>
  left_join(ws_winner, by = c("winner")) |>
  replace_na(list(is_winner_ws = 0))
```


### Use the library

```{r}
library(abdwr3edata)
set.seed(111653)
results_1 <- one_simulation_68(0.20). # 0.20 is the standard deviation of talen
results_1
```

### Display standings function

```{r}
display_standings <- function(data, league) {
  data |> 
    filter(League == league) |>
    select(Team, Wins) |>
    mutate(Losses = 162 - Wins) |>
    arrange(desc(Wins))
}
```

### Simulating many seasons

```{r}
set.seed(111653)
many_results <- rep(0.20, 1000) |>
  map(one_simulation_68) |>
  list_rbind()
```


### Plot it

```{r}
ggplot(many_results, aes(Talent, Wins)) +
  geom_point(alpha = 0.05)
```


### Logistic fit 

```{r}
fit1 <- glm(
  Winner.Lg ~ Talent,
  data = many_results, family = binomial
)
fit2 <- glm(
  Winner.WS ~ Talent,
  data = many_results, family = binomial
)
```

Use `predict` to interpet results.

```{r}
tdf <- tibble(
  Talent =  seq(-0.4, 0.4, length.out = 100)
)
tdf |>
  mutate(
    Pennant = predict(fit1, newdata = tdf, type = "response"),
    `World Series` = predict(fit2, newdata = tdf, type = "response")
  ) |>
  pivot_longer(
    cols = -Talent,
    names_to = "Outcome", 
    values_to = "Probability"
  ) |>
  ggplot(aes(Talent, Probability, color = Outcome)) +
  geom_line() + ylim(0, 1)  
```




